Estimation of Intracortical Connectivity in a Dynamical Neural Field Model

### 2010-03-15 ltx/BrainIDE.tex
some updates to the paper

Even though we will model  $v_0$ as being homogeneous across the cortical sheet we will express it as being spatially dependant allowing for generality when constructing the state-space representation in the next section. The parameters in equation~\ref{ActivationFunction} are detailed in Table 1. By combining Eqs.~\ref{SpikesToPotential} and \ref{RateBasedInteractions} we get the spatiotemporal model
\begin{equation}\label{FullDoubleIntModel}
v\left(r,t\right) = \int\limits_{-\infty}^t h\left(t - t'\right) \left(\int\limits_\Omega  w\left(r,r'\right)f\left( v\left( r',\bar t \right)\right)dr' + e\left(r,t\right)\right) dt'
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{GreensFuncDef}
Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This provides the most general form
\begin{equation}\label{GenForm}
L\left( \frac{\partial }{dt} \right)v\left( {r,t} \right) = \int_\Omega  {w\left( {r,r'} \right)f\left( {v\left( {r',\bar t} \right)} \right)dr'}  + de\left( r,t \right),
\end{equation}
where $L$ is a polynomial of order $n$ with constant coefficients that provides an $n^{th}$ order model. Most studies considering this model use first or second order time derivatives. We shall use the first order model giving
\begin{equation}\label{FinalForm1}
dv\left( {r,t} \right) + \alpha v\left( r,t \right)dt = \alpha\int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',\bar t \right)} \right)dr'dt}  +de\left( r,t \right).
\end{equation}

\section{Discrete Time and Space Model}
We begin the derivation of the state-space model with the continuous time, continuous space IDE neural field model. 
Spatio-temporal interactions can be described by the continuous space continuous time stochastic differential equation
\begin{equation}\label{differential eq}	
dv\left( r,t \right) = \alpha \left( -v\left( r,t \right) + \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t \right)} \right)dr'} \right) dt  + d e\left( r,t \right).
\end{equation}
To simplify the notation, but without loss of generality, \ref{FinalForm1} can be rewritten as
\begin{equation}\label{differential simplified}
	dv\left( r,t \right) = \alpha q\left(v(r,t) \right) dt  + d e\left( r,t \right)
\end{equation}
A one-step Euler method was used to approximate the nonlinear stochastic IDE to a discrete time form. 
\begin{equation}\label{euler}
	v\left(r,t\right) =  v(r,t_0) + \alpha \int_{t_0}^{t} q\left(v(r,\tau) \right) d\tau  + \int_{t_0}^{t} d e\left(r,\tau\right)
\end{equation}
Using an Euler approximation,~\ref{euler} can be approximated at each sampling time by
\begin{equation}\label{euler approx 1}
	v\left(r,t_{k+1} \right) =  v(r,t_k) + \alpha q\left(v(r,t_k)\right) (t_{k+1} - t_k) + (e(r,t_{k+1}) - e(r,t_k))
\end{equation}
which can be rewritten as
\begin{equation}\label{euler approx 2}
	v\left(r,t_k + \delta \right) =  v(r,t_k) + \delta \alpha q\left(v(r,t_k)\right) + e'(r,t_k)
\end{equation}
where $e'(r,t_k)$ is a zero mean Gaussian random variable with covariance $\delta\Sigma_e$. (Note that, because of the way the noise is defined in \ref{differential eq} it is not $\delta^2$ but $\delta$). Expanding $q\left(v(r,t_k)\right)$ we have
\begin{equation}\label{discrete time model}
	v\left(r,t_k + \delta \right) =  v(r,t_k) + \delta \alpha \left( -v\left( r,t_k \right) + \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t_k \right)} \right)dr'} \right) + e'(r,t_k)
\end{equation}
which can be simplified to
\begin{equation}\label{discrete time model2}
	v\left(r,t_k + \delta \right) =  (1 - \delta \alpha) v(r,t_k) + \delta \alpha \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t_k \right)} \right)dr'} + e'(r,t_k)
\end{equation}
Now we define the spatial aspect of the model on a regular square $i,j$ grid of neural masses, where the spatial step size $\Delta \mathbf{r}_i = \Delta \mathbf{r}_j = \Delta $ giving
\begin{equation}\label{discrete space}
	v\left(r_{ij},t_k + \delta \right) =  (1 - \delta \alpha)v(r_{ij},t_k) + \delta \alpha \Delta^2 \sum_{i=-I}^{i=I}{\sum_{j=-J}^{j=J}{\tilde{w}\left( r_{ij},r_{ij}' \right)f\left( v\left( r_{ij}',t_k \right) \right)} }+ e'(r_{i,j},t_k).
\end{equation}
where $\tilde{w}(r_{ij},r_{ij}')$ discretized connectivity kernel. In Eq~\ref{discrete space} we have used periodic boundary conditions to alleviate problems associated with edge effects. The connectivity kernel is in the form of a modified Mexican hat, and is described by the sum of three weighted Gaussian functions as
\begin{equation}
	\tilde{w}(r_{ij},r_{ij}') = \mathbf{\theta}^T e^{\frac{\left(\left(r_i'-r_i\right)\Delta\right)^2}{(2\mathbf{\sigma})^2} + \frac{\left(\left(r_j'-r_j\right)\Delta\right)^2}{(2\mathbf{\sigma})^2}}
\end{equation}
where
\begin{equation}
 \mathbf{\theta}=\begin{bmatrix} \mathbf{\theta}_{1}\\\mathbf{\theta}_{2}\\\mathbf \theta_{3}\end{bmatrix},\quad \mathbf{\sigma}=\begin{bmatrix}\mathbf{\sigma}_1\\\mathbf{\sigma}_2\\ \mathbf{\sigma}_3\end{bmatrix}.
\end{equation}.
The first Gaussian function (parametrized by $\mathbf{\theta}_1, \mathbf{\sigma}_1$) describes local excitation, the second describes lateral inhibition and the third describes medium range excitation.

@@

In order to implement standard estimation techniques we propose a decomposition of the field using a set of Gaussian basis functions. Decomposition allows a continuous field to be represented by a finite dimensional state vector. This allows for the application of standard nonlinear, non-Gaussian state estimation methods such as sequential Monte-Carlo techniques like particle filtering. The field decomposition is described by
\begin{equation}\label{DefFieldDecomp}
{v_t}\left( {\bf{r}} \right) = {\phi ^T}\left( {\bf{r}} \right){x_t}.
\end{equation}
where $\phi(r)$ is the Gaussian basis functions that are scaled by the the state vector. The connectivity kernel can all be decomposed into a set of basis functions
\begin{equation}\label{DefKernelDecomp}
k( \mathbf{r} , \mathbf{r'} ) = \theta^T\psi(\mathbf{r} , \mathbf{r'} )
\end{equation}
Will will assume we know the parametric form of the basis functions, where the parameter $\theta$ is unknown. Each connectivity basis function can be considered a layer in the WC model, representing short range excitation, surround inhibition and long range excitation. Making substitutions we get
\begin{equation}\label{DecompModel}
{\phi ^T}\left( \mathbf{r} \right){x_{t+1}} = T_s\theta^T\int_\Omega  {\psi \left( {{\bf{r}} , {\mathbf{r'}}} \right)f\left( {{\phi ^T}\left( {{\mathbf{r'}}} \right){x_t}} \right)d{\mathbf{r'}}}  - \lambda\phi^T(\mathbf{r})x_t + T_s{e_t}\left( {\mathbf{r}} \right).
\end{equation}
To isolate the state we cross multiply by $\phi(\mathbf{r})$ and integrate over $\mathbf{r}$
\begin{equation}\label{Poo}
\int_\Omega  {\phi \left( {\bf{r}} \right){\phi ^T}\left( {\bf{r}} \right)d{\bf{r}}{x_{t+1}}}  = T_s\int_\Omega  {\phi ( \mathbf{r} )\theta^T\int_\Omega  {\psi \left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}}  - \lambda\int_\Omega \phi(\mathbf{r})\phi^T(\mathbf{r})d\mathbf{r} x_t  + T_s\int_\Omega  {\phi \left( {\mathbf{r}} \right)e_t\left( {\mathbf{r}} \right)d{\mathbf{r}}}.
\end{equation}
Now we define
\begin{equation}\label{DefGamma}
\Gamma  = \int_\Omega  {\phi \left( {\bf{r}} \right){\phi ^T}\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
Substituting equation~\ref{DefGamma} into \ref{Poo} and rearranging gives
\begin{equation}\label{SS Model}
{x_{t+1}} = T_s{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)\theta^T\int_\Omega  {\psi \left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}} - \lambda x_t + T_s{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)e_t\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
We can rearrange to get
\begin{equation}\label{Homogeneous SS Model}
	x_{t+1} = T_s\Gamma ^{ - 1}\int_\Omega  {\phi \left( {\bf{r}} \right)\int_\Omega  {f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)\psi^T \left( {{\bf{r}} - {\bf{r'}}} \right)d{\bf{r'}}} d{\bf{r}}}\theta - \lambda x_t + T_s{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)e_t\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
The disturbance term can be rewritten as 
\begin{equation}
	\Gamma ^{ - 1}\int_\Omega  {\phi \left( \bf{r} \right)e_t\left( \bf{r} \right)d\bf{r}} = \int_\Omega  {\phi \left( \bf{r} \right)\tilde{e}_t\left( \bf{r} \right)d\bf{r}},
\end{equation}
where $\tilde{e}_t(r)~N(0,R'R)$ and $R'$ is the Cholesky decomposition of $\Gamma^{-1}$. This provides the final state-space model
\begin{equation}\label{AbbrevSSModel}
x_{t + 1} = T_sq(x_t)\theta -\lambda x_t + T_s\int_\Omega  {\phi ( \mathbf{r} )\tilde{e}_t( \mathbf{r} )d\mathbf{r}}
\end{equation}
with observation equation
\begin{equation}\label{ObservationEquation}
    y_t( k ) = \int_\Omega  m( k,\mathbf{r} )\phi^T( \mathbf{r} ) d\mathbf{r}x_t  + \varepsilon ( k ),
\end{equation}

>>>>

Even though we will model  $v_0$ as being homogeneous across the cortical sheet we will express it as being spatially dependant allowing for generality when constructing the state-space representation in the next section. The parameters in equation~\ref{ActivationFunction} are detailed in Table 1. By combining Eqs.~\ref{SpikesToPotential} and \ref{RateBasedInteractions} we get the spatiotemporal model
\begin{equation}\label{FullDoubleIntModel}
v\left(r,t\right) = \int\limits_{-\infty}^t h\left(t - t'\right) \left(\int\limits_\Omega  w\left(r,r'\right)f\left( v\left( r',\bar t \right)\right)dr' + e\left(r,t\right)\right) dt'
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{GreensFuncDef}
Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L=\frac{d}{dt} + \alpha$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This provides the most general form
\begin{equation}\label{FinalForm1}
dv\left( {r,t} \right) + \alpha v\left( r,t \right)dt = \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',\bar t \right)} \right)dr'dt}  +de\left( r,t \right).
\end{equation}

\section{Discrete Time and Space Model}
We begin the derivation of the state-space model with the continuous time, continuous space IDE neural field model. 
Spatio-temporal interactions can be described by the continuous space continuous time stochastic differential equation
\begin{equation}\label{differential eq}	
dv\left( r,t \right) =  \left( -\alpha v\left( r,t \right) + \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t \right)} \right)dr'} \right) dt  + d e\left( r,t \right).
\end{equation}
To simplify the notation, but without loss of generality, \ref{FinalForm1} can be rewritten as
\begin{equation}\label{differential simplified}
	dv\left( r,t \right) = q\left(v(r,t) \right) dt  + d e\left( r,t \right)
\end{equation}
A one-step Euler method was used to approximate the nonlinear stochastic IDE to a discrete time form. 
\begin{equation}\label{euler}
	v\left(r,t\right) =  v(r,t_0) + \int_{t_0}^{t} q\left(v(r,\tau) \right) d\tau  + \int_{t_0}^{t} d e\left(r,\tau\right)
\end{equation}
Using an Euler approximation,~\ref{euler} can be approximated at each sampling time by
\begin{equation}\label{euler approx 1}
	v\left(r,t_{k+1} \right) =  v(r,t_k) + q\left(v(r,t_k)\right) (t_{k+1} - t_k) + (e(r,t_{k+1}) - e(r,t_k))
\end{equation}
which can be rewritten as
\begin{equation}\label{euler approx 2}
	v\left(r,t_k + \delta \right) =  v(r,t_k) + \delta q\left(v(r,t_k)\right) + e'(r,t_k)
\end{equation}
where $e'(r,t_k)$ is a zero mean Gaussian random variable with covariance $\delta\Sigma_e$. (Note that, because of the way the noise is defined in \ref{differential eq} it is not $\delta^2$ but $\delta$). Expanding $q\left(v(r,t_k)\right)$ we have
\begin{equation}\label{discrete time model}
	v\left(r,t_k + \delta \right) =  v(r,t_k) + \delta  \left( -\alpha v\left( r,t_k \right) + \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t_k \right)} \right)dr'} \right) + e'(r,t_k)
\end{equation}
which can be simplified to
\begin{equation}\label{discrete time model2}
	v\left(r,t_k + \delta \right) =  (1 - \delta \alpha) v(r,t_k) + \delta \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t_k \right)} \right)dr'} + e'(r,t_k)
\end{equation}
Now we define the spatial aspect of the model on a regular square $i,j$ grid of neural masses, where the spatial step size $\Delta \mathbf{r}_i = \Delta \mathbf{r}_j = \Delta $ giving
\begin{equation}\label{discrete space}
	v\left(r_{ij},t_k + \delta \right) =  (1 - \delta \alpha)v(r_{ij},t_k) + \delta \Delta^2 \sum_{i=-I}^{i=I}{\sum_{j=-J}^{j=J}{\tilde{w}\left( r_{ij},r_{ij}' \right)f\left( v\left( r_{ij}',t_k \right) \right)} }+ e'(r_{i,j},t_k).
\end{equation}
where $\tilde{w}(r_{ij},r_{ij}')$ discretized connectivity kernel. In Eq~\ref{discrete space} we have used periodic boundary conditions to alleviate problems associated with edge effects. The connectivity kernel is in the form of a modified Mexican hat, and is described by the sum of three weighted Gaussian functions as
\begin{equation}
	\tilde{w}(r_{ij},r_{ij}') = \mathbf{\theta}^T e^{\frac{\left(\left(r_i'-r_i\right)\Delta\right)^2}{(2\mathbf{\sigma})^2} + \frac{\left(\left(r_j'-r_j\right)\Delta\right)^2}{(2\mathbf{\sigma})^2}}
\end{equation}
where
\begin{equation}
 \mathbf{\theta}=\begin{bmatrix} \mathbf{\theta}_{1}\\\mathbf{\theta}_{2}\\\mathbf \theta_{3}\end{bmatrix},\quad \mathbf{\sigma}=\begin{bmatrix}\mathbf{\sigma}_1\\\mathbf{\sigma}_2\\ \mathbf{\sigma}_3\end{bmatrix}.
\end{equation}.
The first Gaussian function (parametrized by $\mathbf{\theta}_1, \mathbf{\sigma}_1$) describes local excitation, the second describes lateral inhibition and the third describes medium range excitation.

@@

In order to implement standard estimation techniques we propose a decomposition of the field using a set of Gaussian basis functions. Decomposition allows a continuous field to be represented by a finite dimensional state vector. This allows for the application of standard nonlinear, non-Gaussian state estimation methods such as the unscented Kalman filter. The abbreviate the notation we use the subscript $t$ to index the current time step and $t+1$ for next time step. The field decomposition is described by
\begin{equation}\label{DefFieldDecomp}
v_t\left( \mathbf{r} \right) = \phi ^T\left( \mathbf{r} \right) x_t.
\end{equation}
where $\phi(\mathbf{r})$ is a vector of Gaussian basis functions that are scaled by the state vector. The connectivity kernel can all be decomposed into a set of basis functions
\begin{equation}\label{DefKernelDecomp}
\tilde{w}( \mathbf{r} , \mathbf{r'} ) = \theta^T\psi(\mathbf{r} , \mathbf{r'} )
\end{equation}
Will will assume we know the parametric form of the basis functions, where the parameter $\theta$ is unknown. Each connectivity basis function can be considered a layer in the WC model, representing short range excitation, surround inhibition and mid-range excitation. Making substitutions of~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} into~\ref{discrete space} we get

....up to here....

\begin{equation}\label{DecompModel}
{\phi ^T}\left( \mathbf{r} \right){x_{t+1}} = T_s\theta^T\int_\Omega  {\psi \left( {{\bf{r}} , {\mathbf{r'}}} \right)f\left( {{\phi ^T}\left( {{\mathbf{r'}}} \right){x_t}} \right)d{\mathbf{r'}}}  - \lambda\phi^T(\mathbf{r})x_t + T_s{e_t}\left( {\mathbf{r}} \right).
\end{equation}
To isolate the state we cross multiply by $\phi(\mathbf{r})$ and integrate over $\mathbf{r}$
\begin{equation}\label{Poo}
\int_\Omega  {\phi \left( {\bf{r}} \right){\phi ^T}\left( {\bf{r}} \right)d{\bf{r}}{x_{t+1}}}  = T_s\int_\Omega  {\phi ( \mathbf{r} )\theta^T\int_\Omega  {\psi \left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}}  - \lambda\int_\Omega \phi(\mathbf{r})\phi^T(\mathbf{r})d\mathbf{r} x_t  + T_s\int_\Omega  {\phi \left( {\mathbf{r}} \right)e_t\left( {\mathbf{r}} \right)d{\mathbf{r}}}.
\end{equation}
Now we define
\begin{equation}\label{DefGamma}
\Gamma  = \int_\Omega  {\phi \left( {\bf{r}} \right){\phi ^T}\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
Substituting equation~\ref{DefGamma} into \ref{Poo} and rearranging gives
\begin{equation}\label{SS Model}
{x_{t+1}} = T_s{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)\theta^T\int_\Omega  {\psi \left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}} - \lambda x_t + T_s{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)e_t\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
We can rearrange to get
\begin{equation}\label{Homogeneous SS Model}
	x_{t+1} = T_s\Gamma ^{ - 1}\int_\Omega  {\phi \left( {\bf{r}} \right)\int_\Omega  {f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)\psi^T \left( {{\bf{r}} - {\bf{r'}}} \right)d{\bf{r'}}} d{\bf{r}}}\theta - \lambda x_t + T_s{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)e_t\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
The disturbance term can be rewritten as 
\begin{equation}
	\Gamma ^{ - 1}\int_\Omega  {\phi \left( \bf{r} \right)e_t\left( \bf{r} \right)d\bf{r}} = \int_\Omega  {\phi \left( \bf{r} \right)\tilde{e}_t\left( \bf{r} \right)d\bf{r}},
\end{equation}
where $\tilde{e}_t(r)~N(0,R'R)$ and $R'$ is the Cholesky decomposition of $\Gamma^{-1}$. This provides the final state-space model
\begin{equation}\label{AbbrevSSModel}
x_{t + 1} = T_sq(x_t)\theta -\lambda x_t + T_s\int_\Omega  {\phi ( \mathbf{r} )\tilde{e}_t( \mathbf{r} )d\mathbf{r}}
\end{equation}
with observation equation
\begin{equation}\label{ObservationEquation}
    y_t( k ) = \int_\Omega  m( k,\mathbf{r} )\phi^T( \mathbf{r} ) d\mathbf{r}x_t  + \varepsilon ( k ),
\end{equation}
